Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-28 paper.tex
edits to section 4 with corresponding tweaks to appendix A.  General editing for clarity and conciseness.  Also attempted to blunt perceived criticism of VGSB (which contains many aspects) by focusing on the 'fully radially local approximation'.

%\documentclass[12pt]{aastex}
\documentclass[iop]{emulateapj}
%\documentclass{article}
%\usepackage{draftwatermark}
\usepackage{paralist}
\shorttitle{Vertical shear instability}
\shortauthors{M-K. Lin, A. Youdin}

\usepackage{amsmath}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\w}{ \widetilde{W}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12} 
\defcitealias{nelson13}{N13}
\defcitealias{barker15}{BL15}
\defcitealias{mcnally14}{MP14}
\def \OmK {\Omega_{\rm K}}

@@

A purely hydrodynamic mechanism could circumvent difficulties with non-ideal MHD, but must overcome
the strong centrifugal stability imposed by the positive radial specific angular
momentum gradient in nearly Keplerian disks \citep{balbus96}.
One possible route to hydrodynamic turbulence is the vertical shear
instability (VSI, \citealp{urpin98, urpin03, nelson13}, hereafter \citetalias{nelson13}).  The
basic mechanism of the VSI  in disks was first 
identified in the context of differentially rotating stars \citep[][]{goldreich67,fricke68}.   
The VSI arises when vertical shear, i.e. \ a variation in orbital motion along the rotation axis, 
destabilizes inertial-gravity waves, which are oscillations 
with rotation and buoyancy as restoring forces. Vertical shear
occurs wherever the disk is baroclinic, i.e.\ when constant 
density and constant pressure surfaces are misaligned.  Baroclinicity, and thus vertical shear, is 
practically unavoidable in astrophysical disks, except at special locations like the
midplane.

To overcome centrifugal stabilization, the VSI triggers motions which are vertically elongated 
and radially narrow.  %This geometry minimizes the effects of the radial gradient of angular momentum.
Vertical elongation taps the free energy of the vertical shear \citep{umurhan13}, but is also subject to 
the stabilizing effects of vertical buoyancy if the disk is stably
stratified. To overcome vertical buoyancy, the VSI requires a short cooling time
(\citealp{goldreich67}; \citetalias{nelson13}). Rapid radiative 
cooling, equivalently a short thermal relaxation timescale, brings a
moving fluid element into thermal equilibrium with 
its surroundings, thereby diminishing buoyancy. %not obvious actually
                                %- requires assumption of ideal gas
                                %e.o.s. and pressure eqm established
                                %quickly 
An isothermal equation of state implies instantaneous cooling (and
heating), and is the ideal context for studying the VSI
(\citealp{urpin03}; \citetalias{nelson13}; \citealp{mcnally14}, hereafter \citetalias{mcnally14}; \citealp{barker15},
hereafter \citetalias{barker15}).

>>>>

%\documentclass[12pt]{aastex}
\documentclass[iop, numberedappendix]{emulateapj}
%\documentclass{article}
%\usepackage{draftwatermark}
%\usepackage{paralist}
\shorttitle{Vertical shear instability}
\shortauthors{M-K. Lin, A. Youdin}

\usepackage{amsmath}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\w}{ \widetilde{W}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12} 
\defcitealias{nelson13}{N13}
\defcitealias{barker15}{BL15}
\defcitealias{mcnally14}{MP14}
\defcitealias{goldreich67}{GS67}

\def \OmK {\Omega_{\rm K}}

@@

A purely hydrodynamic mechanism could circumvent difficulties with non-ideal MHD, but must overcome
the strong centrifugal stability imposed by the positive radial specific angular
momentum gradient in nearly Keplerian disks \citep{balbus96}.
One possible route to hydrodynamic turbulence is the vertical shear
instability (VSI, \citealp{urpin98, urpin03, nelson13}, hereafter \citetalias{nelson13}).  The
basic mechanism of the VSI  in disks was first 
identified in the context of differentially rotating stars (\citealp{goldreich67}, hereafter \citetalias{goldreich67}, \citealp{fricke68}).   
The VSI arises when vertical shear, i.e. \ a variation in orbital motion along the rotation axis, 
destabilizes inertial-gravity waves, which are oscillations 
with rotation and buoyancy as restoring forces. Vertical shear
occurs wherever the disk is baroclinic, i.e.\ when constant 
density and constant pressure surfaces are misaligned.  Baroclinicity, and thus vertical shear, is 
practically unavoidable in astrophysical disks, except at special locations like the
midplane.

To overcome centrifugal stabilization, the VSI triggers motions which are vertically elongated 
and radially narrow.  %This geometry minimizes the effects of the radial gradient of angular momentum.
Vertical elongation taps the free energy of the vertical shear \citep{umurhan13}, but is also subject to 
the stabilizing effects of vertical buoyancy if the disk is stably
stratified. To overcome vertical buoyancy, the VSI requires a short cooling time
(\citetalias{goldreich67}; \citetalias{nelson13}). Rapid radiative 
cooling, equivalently a short thermal relaxation timescale, brings a
moving fluid element into thermal equilibrium with 
its surroundings, thereby diminishing buoyancy. %not obvious actually
                                %- requires assumption of ideal gas
                                %e.o.s. and pressure eqm established
                                %quickly 
An isothermal equation of state implies instantaneous cooling (and
heating), and is the ideal context for studying the VSI
(\citealp{urpin03}; \citetalias{nelson13}; \citealp{mcnally14}, hereafter \citetalias{mcnally14}; \citealp{barker15},
hereafter \citetalias{barker15}).
